---
title: "22076699 Assignment"
author: "COMP7024 Programming for Data Science"
output: pdf_document
---
 * Student Surname: Rana
 * Student Firstname: Shobhit
 * Student ID: 22076699
 * Subject Name: Programming for Data Science
 * Subject Code: COMP7024

***
**Declaration**
By including this statement, we the authors of this work, verify that:
• We hold a copy of this assignment that we can produce if the original is lost or damaged.
• We here by certify that no part of this assignment/product has been copied from any otherstudent’s work or from any other source except where due acknowledgement is made in the assignment.
• No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
• We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism (which may retain a copy on its database for future plagiarism checking).
• We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

***

```{r}
setwd("~/Documents/R files")
patientsPG1 = read.csv("patientsPG1.csv")
patientsPG2 = read.csv("patientsPG2.csv")
conditionsPG1 = read.csv("conditionsPG1.csv")
conditionsPG2 = read.csv("conditionsPG2.csv")
encountersPG1 = read.csv("encountersPG1.csv")
encountersPG2 = read.csv("encountersPG2.csv")
library(ggplot2)
library(dplyr)
library(tidyverse)


##Q1
##For dataset 1

#Merging patientsPG1 to conditionsPG1 based on Id and Patient Id columns
patientCondPG1 = merge.data.frame(patientsPG1, conditionsPG1, by.x = "Id", by.y = "PATIENT")

#Finding mean death date for year 2020
patientCondPG1$DEATHDATE <- as.Date(patientCondPG1$DEATHDATE)
patientCondPG1_2020 <- patientCondPG1 %>% filter(format(DEATHDATE, "%Y") == "2020")
mean_date <- mean(patientCondPG1_2020$DEATHDATE, na.rm = TRUE)


#calculating age by first converting date to months then subtracting with mean date or death date and converting floor value to numeric
age_calculator = function(x, y){
  result = floor((as.numeric(y) - as.numeric(x))/365)
  return(as.numeric(result))
}

#Using mean death date in 2020 as the dataset is from 2020 for people still alive
patientCondPG1$Age = ifelse(is.na(patientCondPG1$DEATHDATE), age_calculator(as.Date(patientCondPG1$BIRTHDATE), mean_date), age_calculator(as.Date(patientCondPG1$BIRTHDATE), as.Date(patientCondPG1$DEATHDATE)))

#Making a subset that contains only covid patients
covidPatients = subset(patientCondPG1, patientCondPG1$DESCRIPTION == "COVID-19" | patientCondPG1$DESCRIPTION == "Suspected COVID-19")

#Frequency of covid patients across counties
countyCovid = covidPatients %>%
  count(COUNTY)

# Visualising the distribution of patients with COVID-19 in different counties
ggplot(data = countyCovid) + 
  geom_histogram(mapping = aes(x = COUNTY, y = n), stat = "identity") +
  theme(axis.title.x = element_text(vjust = 1)) +
  labs(title = "Distribution of COVID-19 Patients", x = "County", y = "COVID-19 Cases") +
  theme(axis.title.x = element_text(margin = margin(t=20))) +
  coord_flip()
```

```{r}
#Defining age groups and its corresponding labels
ageGrp <- c(0, 18, 35, 50, Inf)
ageGrpLabel <- c("0-18", "19-35", "36-50", "51+")

#Putting labels for every age that falls under a particular age_group
covidPatients$age_group = cut(covidPatients$Age, breaks = ageGrp, labels = ageGrpLabel)
covidPatients <- covidPatients %>% filter(!is.na(age_group))

#Frequency of patients in various age group
ageCovid = covidPatients %>%
  count(age_group)

#Visualizing age distribution
ggplot(data = ageCovid, aes(x = age_group, y = n)) +
  geom_histogram(stat = "identity") +
  labs(title = "Distribution of COVID-19 patients across Age Groups", x = "Age Groups", y = "COVID-19 Cases")
```






```{r}
##For dataset 2

#Merging patientsPG2 to conditionsPG2 based on Id and Patient Id columns
patientCondPG2 = merge.data.frame(patientsPG2, conditionsPG2, by.x = "Id", by.y = "PATIENT")

#Finding mean death date for year 2020
patientCondPG2$DEATHDATE <- as.Date(patientCondPG2$DEATHDATE)
patientCondPG2_2020 <- patientCondPG2 %>% filter(format(DEATHDATE, "%Y") == "2020")
mean_date1 <- mean(patientCondPG2_2020$DEATHDATE, na.rm = TRUE)

#Using mean death date in 2020 as the dataset is from 2020 for people still alive
patientCondPG2$Age = ifelse(is.na(patientCondPG2$DEATHDATE), age_calculator(as.Date(patientCondPG2$BIRTHDATE), mean_date1), age_calculator(as.Date(patientCondPG2$BIRTHDATE), as.Date(patientCondPG2$DEATHDATE)))

#Making a subset that contains only covid patients
covidPatients1 = subset(patientCondPG2, patientCondPG2$DESCRIPTION == "COVID-19" | patientCondPG2$DESCRIPTION == "Suspected COVID-19")

#Frequency of covid patients across counties
countyCovid1 = covidPatients1 %>%
  count(COUNTY)

# Visualising the distribution of patients with COVID-19 in different counties
ggplot(data = countyCovid1) + 
  geom_histogram(mapping = aes(x = COUNTY, y = n), stat = "identity") +
  theme(axis.title.x = element_text(vjust = 1)) +
  labs(title = "Distribution of COVID-19 Patients", x = "County", y = "COVID-19 Cases") +
  theme(axis.title.x = element_text(margin = margin(t=20))) +
  coord_flip()
```


```{r}
#Defining age groups and its corresponding labels
ageGrp <- c(0, 18, 35, 50, Inf)
ageGrpLabel <- c("0-18", "19-35", "36-50", "51+")

#Putting labels for every age that falls under a particular age_group
covidPatients1$age_group = cut(covidPatients1$Age, breaks = ageGrp, labels = ageGrpLabel)
covidPatients1 <- covidPatients1 %>% filter(!is.na(age_group))

#Frequency of patients in various age group
ageCovid1 = covidPatients1 %>%
  count(age_group)

#Visualizing age distribution
ggplot(data = ageCovid1, aes(x = age_group, y = n)) +
  geom_histogram(stat = "identity") +
  labs(title = "Distribution of COVID-19 patients across Age Groups", x = "Age Groups", y = "COVID-19 Cases")
```


```{r}
##Q2
##For dataset 1
##For both genders
#Counts the frequency of Description except covid-19 and suspected .
symptoms = table(conditionsPG1$DESCRIPTION,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting its data type to data frame
symptoms_df = as.data.frame(symptoms[conditionsPG1$DESCRIPTION])
#Ommiting out Nas and duplicate values 
symptoms_df = distinct(na.omit(symptoms_df))
#Changing the order to descending order of Freq
symptoms_df[order(symptoms_df$Freq, decreasing = TRUE),]
#Showing top 10 values that are descending in order
top10_symptoms = head(symptoms_df[order(symptoms_df$Freq, decreasing = TRUE),], 10)
top10_symptoms
```
```{r}
##For M
#Making subset of data with only Male patients
symptomsM = subset(patientCondPG1, patientCondPG1$GENDER == "M")
#Creating frequency table and excluding covid-19 and suspected cases
symptomsM_tbl = table(c(symptomsM$DESCRIPTION) ,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting to data frame
symptomsM_df = as.data.frame(symptomsM_tbl[symptomsM$DESCRIPTION])
#Adding gender to frequency table
symptomsM_df$GENDER = symptomsM$GENDER
#Ommiting Nas and duplicate values
symptomsM_df = distinct(na.omit(symptomsM_df))
#Printing top 10 symptoms in descending order
top10_symptomsM = head(symptomsM_df[order(symptomsM_df$Freq, decreasing = TRUE),], 10)
top10_symptomsM
```

```{r}
##For F
#Making subset of data with only Female patients
symptomsF = subset(patientCondPG1, patientCondPG1$GENDER == "F")
#Creating frequency table and excluding covid-19 and suspected cases
symptomsF_tbl = table(c(symptomsF$DESCRIPTION) ,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting to data frame
symptomsF_df = as.data.frame(symptomsF_tbl[symptomsF$DESCRIPTION])
#Adding gender in frequency table
symptomsF_df$GENDER = symptomsF$GENDER
#Ommiting Nas and duplicate values
symptomsF_df = distinct(na.omit(symptomsF_df))
#Printing top 10 symptoms in descending order
top10_symptomsF = head(symptomsF_df[order(symptomsF_df$Freq, decreasing = TRUE),], 10)
top10_symptomsF
```

```{r}
##For dataset 2
##For both genders
#Counts the frequency of Description except covid-19 and suspected .
symptoms1 = table(conditionsPG2$DESCRIPTION,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting its data type to data frame
symptoms_df1 = as.data.frame(symptoms1[conditionsPG2$DESCRIPTION])
#Ommiting out Nas and duplicate values 
symptoms_df1 = distinct(na.omit(symptoms_df1))
#Changing the order to descending order of Freq
symptoms_df1[order(symptoms_df1$Freq, decreasing = TRUE),]
#Showing top 10 values that are descending in order
top10_symptoms1 = head(symptoms_df1[order(symptoms_df1$Freq, decreasing = TRUE),], 10)
top10_symptoms1
```

```{r}
##For M
#Making subset of data with only Male patients
symptomsM1 = subset(patientCondPG2, patientCondPG2$GENDER == "M")
#Creating frequency table and excluding covid-19 and suspected cases
symptomsM_tbl1 = table(c(symptomsM1$DESCRIPTION) ,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting to data frame
symptomsM_df1 = as.data.frame(symptomsM_tbl1[symptomsM1$DESCRIPTION])
#Adding gender to frequency table
symptomsM_df1$GENDER = symptomsM1$GENDER
#Ommiting Nas and duplicate values
symptomsM_df1 = distinct(na.omit(symptomsM_df1))
#Printing top 10 symptoms in descending order
top10_symptomsM1 = head(symptomsM_df1[order(symptomsM_df1$Freq, decreasing = TRUE),], 10)
top10_symptomsM1
```

```{r}
##For F
#Making subset of data with only Female patients
symptomsF1 = subset(patientCondPG2, patientCondPG2$GENDER == "F")
#Creating frequency table and excluding covid-19 and suspected cases
symptomsF_tbl1 = table(c(symptomsF1$DESCRIPTION) ,exclude = c("COVID-19", "Suspected COVID-19"))
#Converting to data frame
symptomsF_df1 = as.data.frame(symptomsF_tbl1[symptomsF1$DESCRIPTION])
#Adding gender in frequency table
symptomsF_df1$GENDER = symptomsF1$GENDER
#Ommiting Nas and duplicate values
symptomsF_df1 = distinct(na.omit(symptomsF_df1))
#Printing top 10 symptoms in descending order
top10_symptomsF1 = head(symptomsF_df1[order(symptomsF_df1$Freq, decreasing = TRUE),], 10)
top10_symptomsF1
```

```{r}
##Q3
##For dataset 1
#Merging/Inner join dataset covidPatients and encountersPG1
covidEncounters = merge.data.frame(covidPatients, encountersPG1, by.x = "Id", by.y = "PATIENT")

#Making subset of covidEncounters for the rows that contain ambulatory, emergency, inpatient or urgent care in ENCOUNTERCLASS
hospitalisationRate = subset(covidEncounters, covidEncounters$ENCOUNTERCLASS %in% c("ambulatory", "emergency", "inpatient", "urgent care"))

#Counting for hospitalisation rates for each gender
genderHospitalizationCount = hospitalisationRate %>%
  group_by(GENDER, ENCOUNTERCLASS) %>%
  count() %>%
  arrange(desc(n))

#Visualizing hospitalization rate for each gender
ggplot(data = genderHospitalizationCount) +
  geom_histogram(mapping = aes(x = ENCOUNTERCLASS, y = n, fill = GENDER), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Hospitalization rates for COVID-19 Patients",
       x = "Encounter Class",
       y = "Number of Encounters",
       fill = "Gender")

```


```{r}
#Counting hospitalization rates for each county
countyHospitalizationCount = hospitalisationRate %>%
  group_by(COUNTY, ENCOUNTERCLASS) %>%
  count() %>%
  arrange(desc(n))

#Visualizing hospitalization rate for each county
ggplot(data = countyHospitalizationCount) +
  geom_histogram(mapping = aes(x = ENCOUNTERCLASS, y = n, fill = COUNTY), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Hospitalisation rates for COVID-19 Patients",
       x = "Encounter Class",
       y = "Number of Encounters",
       fill = "County")

```

```{r}
##For dataset 2
#Merging/Inner join dataset covidPatients and encountersPG1
covidEncounters1 = merge.data.frame(covidPatients1, encountersPG2, by.x = "Id", by.y = "PATIENT")

#Making subset of covidEncounters for the rows that contain ambulatory, emergency, inpatient or urgent care in ENCOUNTERCLASS
hospitalisationRate1 = subset(covidEncounters1, covidEncounters1$ENCOUNTERCLASS %in% c("ambulatory", "emergency", "inpatient", "urgent care"))

#Counting for hospitalisation rates for each gender
genderHospitalizationCount1 = hospitalisationRate1 %>%
  group_by(GENDER, ENCOUNTERCLASS) %>%
  count() %>%
  arrange(desc(n))

#Visualizing hospitalization rate for each gender
ggplot(data = genderHospitalizationCount1) +
  geom_histogram(mapping = aes(x = ENCOUNTERCLASS, y = n, fill = GENDER), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Hospitalization rates for COVID-19 Patients",
       x = "Encounter Class",
       y = "Number of Encounters",
       fill = "Gender")
```

```{r}
#Counting hospitalization rates for each county
countyHospitalizationCount1 = hospitalisationRate1 %>%
  group_by(COUNTY, ENCOUNTERCLASS) %>%
  count() %>%
  arrange(desc(n))

#Visualizing hospitalization rate for each county
ggplot(data = countyHospitalizationCount1) +
  geom_histogram(mapping = aes(x = ENCOUNTERCLASS, y = n, fill = COUNTY), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Hospitalisation rates for COVID-19 Patients",
       x = "Encounter Class",
       y = "Number of Encounters",
       fill = "County")
```

```{r}
##Q4
##For dataset 1
#Calculating recovery time and flagging patients based on their recoveries
covidRecovery = covidPatients %>%
  mutate(Start_date = as.Date(START), 
         Stop_date = as.Date(STOP),
         Days_to_recover = as.numeric(difftime(Stop_date, Start_date, units = "days")),
         ## Identify if patient has recovered or not recovered with T/F condition
         Recovered = !is.na(Stop_date))

#Count for each genders that have recovered and not recovered
genderRecoveryCount = covidRecovery %>%
  group_by(GENDER, Recovered) %>%
  count() %>%
  arrange(desc(n))

#Visualizing recovery count for each gender
ggplot(data = genderRecoveryCount) +
  geom_histogram(mapping = aes(x = Recovered, y = n, fill = GENDER), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Recovery rates for COVID-19 Patients",
       x = "Recovered",
       y = "Number of patients",
       fill = "Gender")
```

```{r}
#Count for all ages that have recovered and not recovered
ageRecoveryCount = covidRecovery %>%
  group_by(Age, Recovered) %>%
  count() %>%
  arrange(desc(n))

#Visualizing recovery count for all ages
ggplot(data = ageRecoveryCount) +
  geom_histogram(mapping = aes(x = Recovered, y = n, fill = Age), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Recovery rates for COVID-19 Patients",
       x = "Recovered",
       y = "Number of patients",
       fill = "Age")
```

```{r}
##For dataset 2
#Calculating recovery time and flagging patients based on their recoveries
covidRecovery1 = covidPatients1 %>%
  mutate(Start_date = as.Date(START), 
         Stop_date = as.Date(STOP),
         Days_to_recover = as.numeric(difftime(Stop_date, Start_date, units = "days")),
         ## Identify if patient has recovered or not recovered with T/F condition
         Recovered = !is.na(Stop_date))

#Count for each genders that have recovered and not recovered
genderRecoveryCount1 = covidRecovery1 %>%
  group_by(GENDER, Recovered) %>%
  count() %>%
  arrange(desc(n))

#Visualizing recovery count for each gender
ggplot(data = genderRecoveryCount1) +
  geom_histogram(mapping = aes(x = Recovered, y = n, fill = GENDER), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Recovery rates for COVID-19 Patients",
       x = "Recovered",
       y = "Number of patients",
       fill = "Gender")
```

```{r}
#Count for all ages that have recovered and not recovered
ageRecoveryCount1 = covidRecovery1 %>%
  group_by(Age, Recovered) %>%
  count() %>%
  arrange(desc(n))

#Visualizing recovery count for all ages
ggplot(data = ageRecoveryCount1) +
  geom_histogram(mapping = aes(x = Recovered, y = n, fill = Age), 
                 stat = "identity", position = position_dodge()) +
  labs(title = "Recovery rates for COVID-19 Patients",
       x = "Recovered",
       y = "Number of patients",
       fill = "Age")
```

